Supermartingales and Termination Analysis of Statistical PCF

### 2020-12-29 LICS21/antitone.tex
Sec. 3: Supermartingales - cutting down to 2- pp.

\lo{This is a genuine (and simple) \emph{non-affine recursive program}, i.e., recursive programs that can, during the evaluation of the recursive body, make multiple recursive
calls (of a first-order function) from distinct call sites. 
We can use this example to motivate the confluent semantics, though it is perhaps not the most compelling possible.
Reason: one could use a more restrictive but equally expressive recursion \emph{operator} (as opposed to a $\tY$-\emph{combinator}, which has the status of a constant) to implement general recursion which would make it unnecessary to consider an alternative reduction strategy when applying the partial ranking function theorem. 
E.g.~define letrec-terms, $\mathsf{letrec} \, f^{A \to B} \, x^A . M$, as values, via the formation rule
\[{\Gamma, f: A \to B, x:A  \vdash M : B} \over
\Gamma \vdash \mathsf{letrec} \, f^{A \to B} \, x^A . M : A \to B
\]
with the redex rule 
\[
(\mathsf{letrec} \, f^{A \to B} \, x^A . M^B) \, V \to M[\mathsf{letrec} \, f \, x . M / f, V / x].
\]}
\end{example}

@@

\begin{example}[Non-affine recursion with continuous distribution]
\label{ex:non-affine with continuous distribution}
Consider 
\[
\Xi := \tY \lambda f x . \big(\mathsf{let} \; e = \tsample \; \mathsf{in} \;
\mathsf{if}\boldsymbol{(}e \leq p, x, X\boldsymbol{)}\big)
\]
where
\(
X = \big( f^3(x+1) \oplus_e f^2(x+1) \big) \oplus_{\mathit{sig}(x)} f^2(x+1) 
\)
and $\mathit{sig}(x)$ is the Sigmoid function.

@@

\lo{
DRAFT. Motivation: higher-order recursion. This could be placed in \Cref{sec:intro} or \Cref{sec:related}.
}
There is an obvious source of deterministic (i.e.~non-probabilistic) higher-order functions defined by recursion, viz., \emph{higher-order recursion schemes} (HORS) (see e.g.~\citep{DBLP:conf/lics/Ong06,DBLP:conf/lics/Ong15}).
(Incidentally the (sparse) ranking function method is just as applicable to the termination analysis of \emph{deterministic} SPCF programs.)
Recently \cite{DBLP:conf/lics/KobayashiLG19} have extended HORS to \emph{probabilistic higher-order recursion schemes} (PHORS), which are HORS augmented with probabilistic (binary) branching $\oplus_p$.
As HORS are in essence the $\lambda \tY$-calculus \cite{DBLP:conf/lics/Statman02} (i.e.~pure simply-typed lambda calculus with recursion, generated from a finite base type), PHORS are definable in SPCF:
(order-$n$) PHORS is encodable as (order-$n$) CbN SPCF, but the former is strictly less expressive (because the underlying recursion schemes are not Turing complete). 
Some interesting SPCF terms such as \Cref{ex:non-affine with continuous distribution} %\refExample{complexExample} 
cannot be expressed as PHORS.
A relevant result here is that the AST problem is decidable for order-1 PHORS (by reduction to the PSPACE-hard solvability of finite systems of polynomial equations with real coefficients; see \citep{DBLP:journals/jacm/EtessamiY09}), but undecidable for order-2 PHORS.

>>>>

\lo{This is a genuine (and simple) \emph{non-affine recursive program}, i.e., recursive programs that can, during the evaluation of the recursive body, make multiple recursive
calls (of a first-order function) from distinct call sites. 
We can use this example to motivate the confluent trace semantics, though it is perhaps not the most compelling possible.
Reason: one could use a more restrictive but equally expressive recursion \emph{operator} (as opposed to a $\tY$-\emph{combinator}, which is a constant) to implement general recursion which would make it unnecessary to consider an alternative reduction strategy when applying the partial ranking function theorem. 
E.g.~define letrec-terms, $\mathsf{letrec} \, f^{A \to B} \, x^A . M$, as values, via the formation rule
\[{\Gamma, f: A \to B, x:A  \vdash M : B} \over
\Gamma \vdash \mathsf{letrec} \, f^{A \to B} \, x^A . M : A \to B
\]
with the redex rule 
\[
(\mathsf{letrec} \, f^{A \to B} \, x^A . M^B) \, V \to M[\mathsf{letrec} \, f \, x . M / f, V / x].
\]}
\end{example}

@@

\begin{example}[Non-affine recursion with continuous distribution]
\label{ex:non-affine with continuous distribution}
Consider 
\[
\Xi := \tY \lambda f x . \big(\mathsf{let} \; e = \tsample \; \mathsf{in} \;
\mathsf{if}\boldsymbol{(}e \leq p, x, X\boldsymbol{)}\big)
\]
where
\(
X = \big( f^3(x+1) \oplus_e f^2(x+1) \big) \oplus_{\mathit{sig}(x)} f^2(x+1) 
\)
and $\mathit{sig}(x)$ is the sigmoid function.

@@

\lo{
DRAFT. Motivation: higher-order recursion. This could be placed in \Cref{sec:intro} or \Cref{sec:related}.
}
There is an obvious source of deterministic (i.e.~non-probabilistic) higher-order functions defined by recursion, viz., \emph{higher-order recursion schemes} (HORS) (see e.g.~\citep{DBLP:conf/lics/Ong06,DBLP:conf/lics/Ong15}).
(Incidentally the (sparse) ranking function method is just as applicable to the termination analysis of \emph{deterministic} SPCF programs.)
Recently \cite{DBLP:conf/lics/KobayashiLG19} have extended HORS to \emph{probabilistic higher-order recursion schemes} (PHORS), which are HORS augmented with probabilistic (binary) branching $\oplus_p$.
As HORS are in essence the $\lambda \tY$-calculus \citep{DBLP:conf/lics/Statman02} (i.e.~pure simply-typed lambda calculus with recursion, generated from a finite base type), PHORS are definable in SPCF:
(order-$n$) PHORS is encodable as (order-$n$) (call-by-name) SPCF, but the former is strictly less expressive (because the underlying recursion schemes are not Turing complete). 
Some interesting SPCF terms such as \Cref{ex:non-affine with continuous distribution} %\refExample{complexExample} 
cannot be expressed as PHORS.
A relevant result here is that the AST problem is decidable for order-1 PHORS (by reduction to the PSPACE-hard solvability of finite systems of polynomial equations with real coefficients; see \citep{DBLP:journals/jacm/EtessamiY09}), but undecidable for order-2 PHORS.

### 2020-12-29 LICS21/supermartingales-proofs.tex
Sec. 3: Supermartingales - cutting down to 2- pp.

>>>>

\TechnicalRankingSupermartingale*

First, some notation.

Given an $\omega$-sequence (e.g.~$s \in \entrosp$) and $m \geq 0$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.
For $n \geq 0$, define
\begin{align*}
%M_n(s) &:= \pi_0 (\red^n(M, s))\\
\ndraw{n}{s} &:= |\{k < n \mid \exists E: M_k(s) = E[\tsample]\}|
\end{align*}
so that $\pi_1 (\red^n(M, s)) = %\overbrace{\pi_t( \cdots (\pi_t}^{\ndraw n s}(s))$. 
\pi_t( \cdots (\pi_t (s))$, with $\pi_t$ applied ${\ndraw n s}$ times.
%and the filtration $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
The $\calF_n$-measurability of $M_n$ (and hence of $\#\mathrm{draw}_n$) follows from \citep{DBLP:conf/icfp/BorgstromLGS16}.
%$\#\mathrm{draw}_n$ is a stopping time (bounded by $n$). 
\iffalse
\akr{$\#\mathrm{draw}_n$ is not a stopping time. It doesn't look like you actually use this claim anyway, so it should just be fine to remove, but did you mean something different?} \lo{I agree, and I don't actually use this claim.}
\fi
Take $s \in A \in \calF_n$ with $\ndraw{n}{s} = l$.
For any $s'\in \entrosp$, if $s_{\leq l} = s_{\leq l}'$ then $s' \in A$.
It follows that ${\set{s_{\leq l}} \cdot I^\omega} \subseteq A$.
\lo{NOTE. To fix notation:
\(
\sigma(M_{n}) = \sigma\big(\set{M_{n}^{-1}(\alpha_j, U_{\alpha_j})
\mid \alpha_j\in \mathsf{Sk}_j, U_{\alpha_j} \in \mathcal{B}(\Real^j), j \geq 0}\big)
\)
}
%Take $M \in \Lambda^0$ (closed SPCF terms). Define, for each $n \in \omega$, the random variable $M_n : \entrosp \to \Lambda^0$ by $M_n(s) := \pi_0 (\red^n(M, s))$.

\begin{proof}
We show the non-trivial case of $i = 2$.
First we express 
\begin{equation}
f(M_{n+1})[s \in \mathbf{T}_2] = \sum_{i \in \calI} 
f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])[s \in U_i]
\label{eqn:f 2 n+1}
\end{equation}
where 
\begin{itemize}
\item $\calI$ is a countable indexing set
\item $E_i[\cdot][\tsample] \in \mathsf{Sk}_{j_i}$, and $\sigma_i : \entrosp \to \Real^{j_i}$, and $\rho(s) := \pi_h(\pi_1(\red^n(M, s))) = \pi_h(\pi_t^{l_i}(s)) \in \Real$ 
\item $\set{U_i}_{i \in \calI}$ is a partition of $\mathbf{T}_2$, 
where each $U_i$ is determined by a skeletal term $E_i[\cdot][\tsample]$ and a number (of draws) $l_i \leq n$, satisfying
\(
s \in U_i
\iff
(E_i[\underline{r}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s),
\;
\hbox{for some }r \in \Real^{j_i}.
\)
(We use the (measurable) function $\sigma_i : \entrosp \to \Real^{j_i}$ to skolemise the existentially quantified $r$.)
Equivalently
\[
U_i := \#\mathrm{draw}_n^{-1}(l_i) \cap M_n^{-1}(E_i[\cdot][\tsample], \Real^{j_i}) \in \calF_n.
\]
%\lo{I.e.~$(E_i[\underline{\sigma_i(s)}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s)$ iff $s \in U_i$.}
\end{itemize}

Observe that if $s \in U_i$ then $\set{s_{\leq l_i}} \cdot I^\omega \subseteq U_i$;
in fact $(U_i)_{\leq l_i} \cdot I^\omega = U_i$.
%Moreover, for any $A \in \calF_n$, if $s \in A$ then $\set{s_{\leq l}} \cdot I^\omega \subseteq A$.
This means that for any measurable $g : \entrosp \to \Real_{\geq 0}$, if $g(s)$ only depends on the prefix of $s$ of length $(l_i+1)$, then, writing $\widehat{g} : I^{l_i+1} \to \Real_{\geq 0}$ where $g(s) = \widehat{g}(s_{\leq l_i+1})$, we have: for any $A \in \calF_n$ 
\begin{equation}
\int_{A \cap U_i}  \mu_{\entrosp}(\dif s) \, g(s) = 
\int_{(A \cap U_i)_{\leq l_i+1}} \Leb_{l_i+1}(\dif t) \, \widehat{g}(t)
\label{eqn:truncate}
\end{equation}
%(For simplicity, we will write $g' = g$ in the following.)
%where $A_{\leq m} := \set{s_{\leq m} \mid s \in A}$.

Take $s \in U_i$, and set $l = l_i$.
Plainly $\sigma_i(s)$ depends on $s_{\leq l}$, and $\rho(s)$ depends on $s_{\leq l +1}$.
Take $u \in (U_i)_{\leq l}$.
It then follows from the definition of ranking function that
\[%\textstyle
\int_I \Leb(\dif r) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\underline{r}])) \leq f(E_i[\underline{\widehat{\sigma_i}(u)}][\tsample])).
\]
Take $A \in \calF_n$, and integrating both sides, we get
\begin{align*}
& \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, \int_I \textrm{Leb}(\dif r) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\underline{r}])) \\
& \leq \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\tsample])).
\end{align*}
Since $\Leb_{l+1}$ is the (unique) product measure satisfying $\Leb_{l+1}(V \times B) = \Leb_l(V) \cdot \Leb(B)$, and $(U_i)_{\leq l_i} \cdot I^\omega = U_i$, we have
%it follows from the preceding observation that
\begin{align*}
& \int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\widehat{\sigma_i}(u')}][\underline{\widehat{\rho}(u')}])) \\
& \leq 
\int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\widehat{\sigma_i}(u')}][\tsample]))
\end{align*}
and so, by (\ref{eqn:truncate})
\begin{align}
& \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])) \nonumber \\
& \leq 
\int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\tsample])).
\label{eqn:a u ui}
\end{align}

Now, integrating both sides of (\ref{eqn:f 2 n+1}), we have
\begin{align*}
& \int_A \mu_{\entrosp}(\dif s) \, f_{n+1}(M)[s \in \mathbf{T}_i] \\
&= 
\int_A \mu_{\entrosp}(\dif s) \, \sum_{i \in \calI} f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])[s \in U_i] \\
&= 
\sum_{i \in \calI} \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])\\
&\leq 
\sum_{i \in \calI} \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\tsample])
\quad \hbox{$\because$ (\ref{eqn:a u ui})}\\
&= 
\int_{A} \mu_{\entrosp}(\dif s) \, \sum_{i \in \calI} \, f(E_i[\underline{\sigma_i(s)}][\tsample])[s \in U_i]\\
&= \int_{A} \mu_{\entrosp}(\dif s) f_{n}(M)[s \in \mathbf{T}_2]
%\quad \hbox{$\because$ (\ref{eqn:f 2 n+1})}
\end{align*}

\end{proof}
